---
title: "Practical Machine Learning Assignment"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

Here is the project brief:

>>>Background

>>>Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE, echo = FALSE)
library(tidyverse)
library(tidymodels)
library(parsnip)
library(skimr)
library(moments) # for skewness test

train_data <- read_csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
test_data <- read_csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

## Exploration

The first task is to explore the data, to reveal what type of modelling might be appropriate, and what sort of preprocessing is necessary. 

_Note:_ I have elected to use the `tidymodels` package, as it is the natural evolution of the caret package, utilising the `parsnip` package by the same creator, Max Kuhn.  

* Initial view of the dataset shows that it is quite a complex dataset - 160 variables! Definitely worth looking at the background data, which I will comment on next.  
* The __"Class"__ is the outcome here - we have five outcomes ranging from A to E, A being optimum activity.  
>>> Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).  
* the variables will need to be trimmed to those which denote measurements given by the sensors being used (belt, arm, dumbbell).  
* there seem to be a lot of NA values, with only around 100 variables having only 2% of datapoints recorded - I am going to discount these observations as an equipment failure - no other method of imputation could succesfully complete these variables. 
* removing the mostly incomplete predictors and admin variables (timestamp etc) leaves us with 52 predictors.  
* the scale over all of the predictors range from negative -1000 to positive 1000, with some predictors in the 10e-02 scale. There might be a need to scale and centre?  
* initial look at the mini-histograms via `skim` does not show any major skewing errors, but some more investigation of this may be necessary.. About a quarter are highly skewed, another quarter are moderately skewed, and half are roughly symmetrical. 


```{r explore}
head(train_data)
train_skim <- skim(train_data) # inspect variable type, completeness and distribution

train_skim %>% 
  filter(complete_rate > 0.03) %>% 
  select(skim_variable) -> hold_vars # create the predictors that we will keep

train_data %>%
  select(hold_vars$skim_variable) %>% 
  select(-c(user_name, X1, cvtd_timestamp, num_window, new_window, raw_timestamp_part_1, raw_timestamp_part_2)) -> qc1_train_data

# are the classes equally represented?
qc1_train_data %>% 
  count(classe) # yes, roughly speaking

# what is the skewness like in the predictors?
qc1_train_data %>% 
  select(-classe) %>% 
  map_dbl(skewness) -> skewness


skewness %>% 
 as_tibble() %>% 
  mutate(skew_type = case_when(
    abs(value) > 1 ~ "highly",
    abs(value) > 0.5 ~ "moderate",
    abs(value) > 0 ~ "slightly"
  )) -> skewness_rating

skewness_rating$var <- names(skewness)
skewness_rating %>% 
  count(skew_type)
```

## Modelling

Let's attempt to fit three types of models naively, and see what type of results we get. First we will need to subset the data, in order to allow us to perform some vaildation on the data with our models. We will create an 80:20 split, and use the _classe_ variable to stratify.

We will utilise the `recipes` package to create out pre-processing workflow on our training data, which can then be applied to the testing data.

```{r modelling}
exercise_data <- qc1_train_data %>% 
  initial_split(prop = 4/5, strata = classe) # note we're using the subset of 'good' predictors

exercise_data %>% 
  training() %>%
  glimpse() # 15700 x 53


```





